# (PART) Layers {-}


```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE
)
```

# Individual geoms

```{r, message = FALSE}
library(ggplot2)
```


## Basic plot types

**Q1:** What geoms would you use to draw each of the following named plots?

1. Scatterplot
1. Line chart
1. Histogram
1. Bar chart
1. Pie chart

**A:** 

1. Scatterplot: `geom_point()`

1. Line chart:  `geom_line()`

1. Histogram:   `geom_histogram()`

1. Bar chart:   `geom_bar()`

1. Pie chart:   `geom_bar() + coord_polar()`


**Q2:** What’s the difference between `geom_path()` and `geom_polygon()`? What’s the difference between `geom_path()` and `geom_line()`?

**A:** `geom_polygon()` is very similar to `geom_path()` except that the start and end points are connected and the inside is coloured by fill.

`geom_path()` connects the observations in the order in which they appear in the data, but `geom_line()` connects them in order of the variable on the x-axis.



**Q3:** What low-level geoms are used to draw `geom_smooth()`? What about `geom_boxplot()` and `geom_violin()`?

**A:** 

1. `geom_smooth()`: `geom_path()`, and `geom_area()`.

1. `geom_boxplot()`: `geom_line()`, `geom_point()`, and `geom_rect()`.

1. `geom_violin()`: `geom_area()`, and `geom_path()`.

